---
title: "Using the climateR and AOI package"
author: "Kenneth Pomeyie, Scout Jarman, Paul Gaona-Partida"
output:
  pdf_document: default
---

### Introduction
A climatic data element is a measured parameter that helps to specify the climate information of a specific location or region, such as precipitation, temperature, wind speed, and humidity. Raster data models(files) are one way of storing climate data in pixels. The size of these raster files tends to get large as dimensionality increases with time, multiple variables, and multiple models. In most cases, only a small amount of information is needed from these files. The hard way to extract site-specific climate information from raster files involves downloading the files and extracting the required information individually to reduce size. However, this data extraction method is inefficient and memory-limited because R loads everything in RAM. This vignette shows how the climateR and AOI packages can be used to call specific climate data more efficiently.

#### Introduction to AOI and climateR package
Scout


### Advantages of  AOI and climateR package
Paul

### AOI Usage
 Paul and Scout start
### 1. Load libraries
```{r }
# INSTALL
#remotes::install_github("mikejohnson51/AOI")
#remotes::install_github("mikejohnson51/climateR")

#load library
library(sf)
library(climateR)
library(AOI)
library(ggplot2)
library(leaflet.extras)
```

<br>


### 2. Geocode locations 
```{r}
# get point geometry of Logan
logan <- geocode(
  location = "Logan",
  zipcode = NULL, event = NULL, pt = TRUE, bb = FALSE
)
```


```{r}
# check if the location is right
aoi_map(AOI = logan, returnMap = TRUE)
```


```{r}
# use correct location name
logan <- geocode(location = "Logan UT", pt = TRUE)
```


```{r}
# check if the location is right
aoi_map(AOI = logan, returnMap = TRUE)
```


```{r}
# Interactively draw an Area of Interest
aoi_draw()

test <- sf::st_read("aoi_draw/test.gpkg")

aoi_map(test, returnMap = TRUE)

```


<br>

### 3. Queries for state and county boundaries
```{r}
# get Cache county boundaries
cache_aoi <- aoi_get(state = "UT", county = "Cache")
cache_aoi
```


```{r}
# map area of interest
aoi_map(AOI = cache_aoi, returnMap = TRUE)
```

```{r}
# get Utah state boundaries
ut_aoi <- aoi_get(state = "UT", county = "all")
```


```{r}
# map area of interest
aoi_map(AOI = ut_aoi, returnMap = TRUE)
```

<br>

Paul and Scout end

### climateR Usage
### 4. Point Based Time Series (1 site , 1 variable) 1D
```{r}
# get minimum temperature from PRSIM
logan_temp <- climateR::getPRISM(
  AOI = AOI::geocode(location = "Logan UT", pt = TRUE),
  param = "tmin", startDate = "2021-01-01",
  endDate = "2021-12-31"
)
head(logan_temp)
```




<br>

### 5. Multi Point Time Series (3 sites, 2 variables)
```{r}
# get point of 3 sites
sites <- AOI::geocode(location = c(
  "Logan UT", "Salt Lake UT", "Brigham UT"
), pt = TRUE)

```

```{r}
# get precipitation and min temp from GridMET
ut_ppt_tmin <- getGridMET(
  AOI = sites,
  param = c("prcp", "tmin"), startDate = "2021-01-01",
  endDate = "2021-01-31"
)
```

```{r}
# extract precipitation and temperature for specified sites from the raster stack
site_extract <- extract_sites(ut_ppt_tmin, sites, "request")

head(site_extract)
```


<br>

### 6. Multi-layer raster (2 variables, 1 day) 2D
```{r eval=FALSE, include=FALSE}
# get max and min temp  rasters from GridMET
grids <- getGridMET(aoi_get(state = "UT"),
  param = c("tmax", "tmin"),
  startDate = "2000-01-01"
)
plot(raster::stack(grids), main = c("Min Temperature", "Max Temperture"))
```

<br>

### 7. Multi-Cube 4D (3 variables, 3 day)
```{r eval=FALSE, include=FALSE}
# get max and min temp  rasters from GridMET for multi-days
grids <- getGridMET(aoi_get(state = "UT"),
  param = c("tmax", "prcp", "srad"),
  startDate = "2018-10-10",
  endDate = "2018-10-11"
)

plot(raster::stack(grids))
```


